Neural correlates of the uncanny valley effect\\ for robots and hyper-realistic masks

### 2024-07-01 manuscript.Rmd
Finished implementing co-author comments

One previous study by @Saygin2012 measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects^[Repetition suppression is a phenomenon in which the neural responses to repeated presentations of identical or similar stimuli is reduced relative to the response on the first presentation.] in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study [@Urgen2018] identified a difference in the N400 component^[The N400 is an electromagnetic brain potential obtained 400ms after stimulus onset, typically over centro-parietal electrodes. It has been proposed to reflect the extent to which the stimulus presented was surprising or unexpected.] between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in the temporo-parietal cortex, consistent with the fMRI results [@Saygin2012]. The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

@@

All stimuli were displayed on a ViewPixx display running at 120Hz, controlled by an Apple Macintosh computer. The display was gamma corrected using a photometer to ensure that the luminance output was linear. EEG data were collected using a 64-channel Waveguard cap and an ANT Neuroscan system, sampling at 1kHz. Low latency digital triggers were sent between the display and the EEG amplifier using an 8-bit parallel cable.

@@

```{r RobotsData, fig.cap="Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue), android (green) and robot (red) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95 percent confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95 percent confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r RobotsMVPA, fig.cap="Univariate and multivariate comparisons across image type. Panel (a) shows the ERPs comparing human (blue) and robot (red) face images, and panel (b) compares human (blue) and android (green) faces. Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50 percent correct; c,d).", fig.align="center", echo=FALSE}

@@

We also conducted multivariate pattern analysis independently at each time-point for the same two comparisons. The evoked responses for human and robot images caused sufficiently distinct patterns of voltages across the scalp that the pattern classifier could distinguish them from around 100 ms following stimulus onset, with accuracy up to `r round(100*max(meanMVPA[2,]),digits=0)`\% correct (see Figure \ref{fig:RobotsMVPA}c). Bayes factors exceeded 30 for much of the time window between 100 and 800 ms, indicating that the decoding was meaningfully above chance performance (50\% correct). It was also possible to classify between human and android images (see Figure \ref{fig:RobotsMVPA}d), however performance was much poorer, with a maximum of `r round(100*max(meanMVPA[3,]),digits=0)`\% correct. Classification accuracy had an initial peak around 100ms that provided compelling evidence for above chance classification (BF>30), and a later region of above-chance classification between 500 and 700ms. In the Discussion we speculate that these two time periods might correspond to distinct types of signal associated with the uncanny valley. However we first sought to generalise our results to a different stimulus set, and next report the results of Experiment 2 which used hyper-realistic silicone masks.

Finally, we analysed the rating data from a set of 10 questionnaire items for 8 stimuli from each category. The results are summarised in Figure \@ref(fig:robotratings), and in general show differences between stimulus categories along most dimensions. Of particular note, the U-shaped function predicted by the uncanny valley effect was apparent for ratings along the Dislike-Like (Figure \@ref(fig:robotratings)d), Unfriendly-Friendly (Figure \@ref(fig:robotratings)e), and Anxious-Relaxed (Figure \@ref(fig:robotratings)i) dimensions. These are all dimensions with emotional valence, indicating support for the 'uncanniness' of our android stimuli. However we note that the android and robot categories were typically rated as being more similar to each other than to the real human faces.

@@

```{r MasksData, fig.cap="Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue), silicone masks (green) and Halloween masks (red). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95 percent confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95 percent confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r MasksMVPA, fig.cap="Univariate and multivariate comparisons across image type for Experiment 2. Panel (a) shows the ERPs comparing human faces (blue) and Halloween masks (red), and panel (b) compares human faces (blue) and silicone masks (green). Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50 percent correct; c,d).", fig.align="center", echo=FALSE}

@@

We subsequently obtained ratings from an independent sample of 20 participants using the same stimuli as in the main experiment. This time, we asked for explicit ratings of emotional expressiveness, realism, and uncanniness, as well as repeating the binary real face vs mask rating. Real faces were rated highest for emotional expressiveness (M=`r round(mean(averageQdata[,3,1]),digits=1)`) and realism (M=`r round(mean(averageQdata[,3,2]),digits=1)`), and lowest for uncanniness (M=`r round(mean(averageQdata[,3,3]),digits=1)`) (Figure \@ref(fig:maskratings)a-c). The realistic silicone masks were rated highest for uncanniness (M=`r round(mean(averageQdata[,2,3]),digits=1)`), however this was not dramatically higher than for the Halloween masks (M=`r round(mean(averageQdata[,1,3]),digits=1)`). Arguably making judgements of uncanniness is less appropriate for masks that are not intended to be realistic, though our data do qualitatively conform to the U-shaped function expected by the uncanny valley hypothesis. The pattern of d-prime scores (Figure \@ref(fig:maskratings)d) was similar to those obtained in the main experiment (Figure \@ref(fig:MasksData)b), with generally higher scores attributable to the unlimited inspection time permitted here.

@@

Across two experiments using diverse stimuli, we identified a potential neurophysiological signature of the 'uncanny valley' effect. EEG responses to androids or silicone masks could be distinguished from responses to human faces at around 100ms after stimulus onset, and also in a later time window around 500-800ms after stimulus onset. There were no clear differences in the unimodal ERP response at posterior electrodes, but performance of a multivariate pattern classifier was above chance in these time windows. This is a different pattern from that observed for more obviously non-human stimuli (robots and Halloween masks), where there were both univariate and multivariate differences, and the multivariate discrimination accuracy was above chance for an extended time window. Perceptual judgements indicated that identification performance for uncanny valley stimuli was relatively poor, indicating confusion with real human images. The similarity in results across our two experiments is striking and constitutes an internal conceptual replication of our main findings.

@@

Our use of hyper-realistic silicone masks is novel in the context of the study of uncanny valley effects. Previous studies using these masks have demonstrated that they are difficult to distinguish from real faces [@Sanders2017; @Sanders2019], including in applied settings such as simulated passport control [@Robertson2020; @Robertson2024], and show large individual differences [@Sanders2018]. Image analysis indicates that good identification performance of silicone masks is typically based on attention to the region below the eyes [@Sanders2018], however it is plausible that many observers are not explicitly aware of the cues they use to perform this judgement. This might contribute to both the early and late components identified in this study, and presumably also to the subjective sensation of 'uncanniness' that is characteristic of the phenomenon.

>>>>

One previous study by @Saygin2012 measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects^[Repetition suppression is a phenomenon in which the neural response to repeated presentations of identical or similar stimuli is reduced relative to the response on the first presentation.] in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study [@Urgen2018] identified a difference in the N400 component^[The N400 is an electromagnetic brain potential obtained 400ms after stimulus onset, typically over centro-parietal electrodes. It has been proposed to reflect the extent to which the stimulus presented was surprising or unexpected.] between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in the temporo-parietal cortex, consistent with the fMRI results [@Saygin2012]. The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

@@

All stimuli were displayed on a ViewPixx display running at 120Hz, controlled by an Apple Macintosh computer. The display was gamma corrected using a photometer to ensure that the luminance output was linear. EEG data were collected using a 64-channel Waveguard cap and an ANT Neuroscan system, sampling at 1kHz. The ground electrode was located at position $AFz$, and all signals were referenced to the whole head average. Low latency digital triggers were sent between the display and the EEG amplifier using an 8-bit parallel cable.

@@

```{r RobotsData, fig.cap="Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue), android (green) and robot (red) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r RobotsMVPA, fig.cap="Univariate and multivariate comparisons across image type. Panel (a) shows the ERPs comparing human (blue) and robot (red) face images, and panel (b) compares human (blue) and android (green) faces. Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50\\% correct; c,d).", fig.align="center", echo=FALSE}

@@

We also conducted multivariate pattern analysis independently at each time-point for the same two comparisons. The evoked responses for human and robot images caused sufficiently distinct patterns of voltages across the scalp that the pattern classifier could distinguish them from around 100 ms following stimulus onset, with accuracy up to `r round(100*max(meanMVPA[2,]),digits=0)`\% correct (see Figure \ref{fig:RobotsMVPA}c). Bayes factors exceeded 30 for much of the time window between 100 and 800 ms, indicating that the decoding was meaningfully above chance performance (50\% correct). It was also possible to classify between human and android images (see Figure \ref{fig:RobotsMVPA}d), however performance was much poorer, with a maximum of `r round(100*max(meanMVPA[3,]),digits=0)`\% correct. Classification accuracy had an initial peak around 100ms that provided compelling evidence for above chance classification (BF>30), and a later region of above-chance classification between 500 and 700ms. In the Discussion we speculate that these two time periods might correspond to distinct types of signal associated with the uncanny valley.

Finally, we analysed the rating data from a set of 10 questionnaire items for 8 stimuli from each category. The results are summarised in Figure \@ref(fig:robotratings), and in general show differences between stimulus categories along most dimensions. Of particular note, the U-shaped function predicted by the uncanny valley effect was apparent for ratings along the Dislike-Like (Figure \@ref(fig:robotratings)d), Unfriendly-Friendly (Figure \@ref(fig:robotratings)e), and Anxious-Relaxed (Figure \@ref(fig:robotratings)i) dimensions. These are all dimensions with emotional valence, indicating support for the 'uncanniness' of our android stimuli. However we note that the android and robot categories were typically rated as being more similar to each other than to the real human faces. Following Experiment 1, we sought to generalise our results to a different stimulus set, and next report the results of Experiment 2 which used hyper-realistic silicone masks.

@@

```{r MasksData, fig.cap="Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue), silicone masks (green) and Halloween masks (red). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.", fig.align="center", echo=FALSE}

@@

```{r MasksMVPA, fig.cap="Univariate and multivariate comparisons across image type for Experiment 2. Panel (a) shows the ERPs comparing human faces (blue) and Halloween masks (red), and panel (b) compares human faces (blue) and silicone masks (green). Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50\\% correct; c,d).", fig.align="center", echo=FALSE}

@@

We subsequently obtained ratings from an independent sample of 20 participants using the same stimuli as in the EEG experiment. This time, we asked for explicit ratings of emotional expressiveness, realism, and uncanniness, as well as repeating the binary real face vs mask rating. Real faces were rated highest for emotional expressiveness (M=`r round(mean(averageQdata[,3,1]),digits=1)`) and realism (M=`r round(mean(averageQdata[,3,2]),digits=1)`), and lowest for uncanniness (M=`r round(mean(averageQdata[,3,3]),digits=1)`) (Figure \@ref(fig:maskratings)a-c). The realistic silicone masks were rated highest for uncanniness (M=`r round(mean(averageQdata[,2,3]),digits=1)`), however this was not dramatically higher than for the Halloween masks (M=`r round(mean(averageQdata[,1,3]),digits=1)`). Arguably making judgements of uncanniness is less appropriate for masks that are not intended to be realistic, though our data do qualitatively conform to the U-shaped function expected by the uncanny valley hypothesis. The pattern of d-prime scores (Figure \@ref(fig:maskratings)d) was similar to those obtained in the main experiment (Figure \@ref(fig:MasksData)b), with generally higher scores attributable to the unlimited inspection time permitted in this online follow-up experiment

@@

Across two experiments using diverse stimuli, we identified a potential neurophysiological signature of the 'uncanny valley' effect. EEG responses to androids or silicone masks could be distinguished from responses to human faces at around 100ms after stimulus onset, and also in a later time window around 500-800ms after stimulus onset. There were no clear differences in the unimodal ERP response at posterior electrodes, but performance of a multivariate pattern classifier was above chance in these time windows. This is a different pattern from that observed for more obviously non-human stimuli (robots and Halloween masks), where there were both univariate and multivariate differences, and the multivariate discrimination accuracy was above chance for an extended time window. Perceptual judgements indicated that identification performance for uncanny valley stimuli was relatively poor, indicating confusion with real human images. We also confirmed that android images were perceived more negatively than either human or robot images, and that silicone masks were perceived as more uncanny than human faces. The similarity in results across our two experiments is striking and constitutes an internal conceptual replication of our main findings.

@@

Our use of hyper-realistic silicone masks is novel in the context of the study of uncanny valley effects. Previous studies using these masks have demonstrated that they are difficult to distinguish from real faces [@Sanders2017; @Sanders2019], including in applied settings such as simulated passport control [@Robertson2020; @Robertson2024], and show large individual differences [@Sanders2018]. Image analysis indicates that good identification performance for silicone masks is typically based on attention to the region below the eyes [@Sanders2018], however it is plausible that many observers are not explicitly aware of the cues they use to perform this judgement. This might contribute to both the early and late components identified in this study, and presumably also to the subjective sensation of 'uncanniness' that is characteristic of the phenomenon.

### 2024-07-01 manuscript.tex
Finished implementing co-author comments

One previous study by Saygin et al. (2012) measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects\footnote{Repetition suppression is a phenomenon in which the neural responses to repeated presentations of identical or similar stimuli is reduced relative to the response on the first presentation.} in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study (Urgen et al., 2018) identified a difference in the N400 component\footnote{The N400 is an electromagnetic brain potential obtained 400ms after stimulus onset, typically over centro-parietal electrodes. It has been proposed to reflect the extent to which the stimulus presented was surprising or unexpected.} between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in the temporo-parietal cortex, consistent with the fMRI results (Saygin et al., 2012). The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

@@

All stimuli were displayed on a ViewPixx display running at 120Hz, controlled by an Apple Macintosh computer. The display was gamma corrected using a photometer to ensure that the luminance output was linear. EEG data were collected using a 64-channel Waveguard cap and an ANT Neuroscan system, sampling at 1kHz. Low latency digital triggers were sent between the display and the EEG amplifier using an 8-bit parallel cable.

@@

\caption{Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue), android (green) and robot (red) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95 percent confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95 percent confidence interval, and the grey rectangle at the foot indicates the stimulus duration.}\label{fig:RobotsData}
\end{figure}

@@

\caption{Univariate and multivariate comparisons across image type. Panel (a) shows the ERPs comparing human (blue) and robot (red) face images, and panel (b) compares human (blue) and android (green) faces. Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50 percent correct; c,d).}\label{fig:RobotsMVPA}
\end{figure}

We also conducted multivariate pattern analysis independently at each time-point for the same two comparisons. The evoked responses for human and robot images caused sufficiently distinct patterns of voltages across the scalp that the pattern classifier could distinguish them from around 100 ms following stimulus onset, with accuracy up to 72\% correct (see Figure \ref{fig:RobotsMVPA}c). Bayes factors exceeded 30 for much of the time window between 100 and 800 ms, indicating that the decoding was meaningfully above chance performance (50\% correct). It was also possible to classify between human and android images (see Figure \ref{fig:RobotsMVPA}d), however performance was much poorer, with a maximum of 62\% correct. Classification accuracy had an initial peak around 100ms that provided compelling evidence for above chance classification (BF\textgreater30), and a later region of above-chance classification between 500 and 700ms. In the Discussion we speculate that these two time periods might correspond to distinct types of signal associated with the uncanny valley. However we first sought to generalise our results to a different stimulus set, and next report the results of Experiment 2 which used hyper-realistic silicone masks.

Finally, we analysed the rating data from a set of 10 questionnaire items for 8 stimuli from each category. The results are summarised in Figure \ref{fig:robotratings}, and in general show differences between stimulus categories along most dimensions. Of particular note, the U-shaped function predicted by the uncanny valley effect was apparent for ratings along the Dislike-Like (Figure \ref{fig:robotratings}d), Unfriendly-Friendly (Figure \ref{fig:robotratings}e), and Anxious-Relaxed (Figure \ref{fig:robotratings}i) dimensions. These are all dimensions with emotional valence, indicating support for the `uncanniness' of our android stimuli. However we note that the android and robot categories were typically rated as being more similar to each other than to the real human faces.

@@

\caption{Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue), silicone masks (green) and Halloween masks (red). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95 percent confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95 percent confidence interval, and the grey rectangle at the foot indicates the stimulus duration.}\label{fig:MasksData}
\end{figure}

@@

\caption{Univariate and multivariate comparisons across image type for Experiment 2. Panel (a) shows the ERPs comparing human faces (blue) and Halloween masks (red), and panel (b) compares human faces (blue) and silicone masks (green). Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50 percent correct; c,d).}\label{fig:MasksMVPA}
\end{figure}

@@

We subsequently obtained ratings from an independent sample of 20 participants using the same stimuli as in the main experiment. This time, we asked for explicit ratings of emotional expressiveness, realism, and uncanniness, as well as repeating the binary real face vs mask rating. Real faces were rated highest for emotional expressiveness (M=3.9) and realism (M=5.9), and lowest for uncanniness (M=2.7) (Figure \ref{fig:maskratings}a-c). The realistic silicone masks were rated highest for uncanniness (M=4.5), however this was not dramatically higher than for the Halloween masks (M=4.3). Arguably making judgements of uncanniness is less appropriate for masks that are not intended to be realistic, though our data do qualitatively conform to the U-shaped function expected by the uncanny valley hypothesis. The pattern of d-prime scores (Figure \ref{fig:maskratings}d) was similar to those obtained in the main experiment (Figure \ref{fig:MasksData}b), with generally higher scores attributable to the unlimited inspection time permitted here.

@@

Across two experiments using diverse stimuli, we identified a potential neurophysiological signature of the `uncanny valley' effect. EEG responses to androids or silicone masks could be distinguished from responses to human faces at around 100ms after stimulus onset, and also in a later time window around 500-800ms after stimulus onset. There were no clear differences in the unimodal ERP response at posterior electrodes, but performance of a multivariate pattern classifier was above chance in these time windows. This is a different pattern from that observed for more obviously non-human stimuli (robots and Halloween masks), where there were both univariate and multivariate differences, and the multivariate discrimination accuracy was above chance for an extended time window. Perceptual judgements indicated that identification performance for uncanny valley stimuli was relatively poor, indicating confusion with real human images. The similarity in results across our two experiments is striking and constitutes an internal conceptual replication of our main findings.

@@

Our use of hyper-realistic silicone masks is novel in the context of the study of uncanny valley effects. Previous studies using these masks have demonstrated that they are difficult to distinguish from real faces (Sanders et al., 2017, 2019), including in applied settings such as simulated passport control (Robertson et al., 2020, 2024), and show large individual differences (Sanders \& Jenkins, 2018). Image analysis indicates that good identification performance of silicone masks is typically based on attention to the region below the eyes (Sanders \& Jenkins, 2018), however it is plausible that many observers are not explicitly aware of the cues they use to perform this judgement. This might contribute to both the early and late components identified in this study, and presumably also to the subjective sensation of `uncanniness' that is characteristic of the phenomenon.

>>>>

One previous study by Saygin et al. (2012) measured fMRI responses to moving stimuli designed to elicit an uncanny valley effect. They found repetition suppression effects\footnote{Repetition suppression is a phenomenon in which the neural response to repeated presentations of identical or similar stimuli is reduced relative to the response on the first presentation.} in action-specific brain regions responding to movies of androids that had a biological appearance, but mechanical motion. These effects were stronger than for movies of humans or mechanical robots performing the same actions. A more recent electroencephalography (EEG) study (Urgen et al., 2018) identified a difference in the N400 component\footnote{The N400 is an electromagnetic brain potential obtained 400ms after stimulus onset, typically over centro-parietal electrodes. It has been proposed to reflect the extent to which the stimulus presented was surprising or unexpected.} between dynamic and static conditions using the same stimuli. Although this difference was strongest over frontal electrodes, source reconstruction of the N400 itself suggested a left-lateralised source in the temporo-parietal cortex, consistent with the fMRI results (Saygin et al., 2012). The authors interpret both of these findings as being due to the discrepancy between the human-like appearance and the clearly non-biological motion of the robot.

@@

All stimuli were displayed on a ViewPixx display running at 120Hz, controlled by an Apple Macintosh computer. The display was gamma corrected using a photometer to ensure that the luminance output was linear. EEG data were collected using a 64-channel Waveguard cap and an ANT Neuroscan system, sampling at 1kHz. The ground electrode was located at position \(AFz\), and all signals were referenced to the whole head average. Low latency digital triggers were sent between the display and the EEG amplifier using an 8-bit parallel cable.

@@

\caption{Summary of response data and grand mean ERP for Experiment 1. Panel (a) shows d-prime scores for identifying images of human (blue), android (green) and robot (red) faces. Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.}\label{fig:RobotsData}
\end{figure}

@@

\caption{Univariate and multivariate comparisons across image type. Panel (a) shows the ERPs comparing human (blue) and robot (red) face images, and panel (b) compares human (blue) and android (green) faces. Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50\% correct; c,d).}\label{fig:RobotsMVPA}
\end{figure}

We also conducted multivariate pattern analysis independently at each time-point for the same two comparisons. The evoked responses for human and robot images caused sufficiently distinct patterns of voltages across the scalp that the pattern classifier could distinguish them from around 100 ms following stimulus onset, with accuracy up to 72\% correct (see Figure \ref{fig:RobotsMVPA}c). Bayes factors exceeded 30 for much of the time window between 100 and 800 ms, indicating that the decoding was meaningfully above chance performance (50\% correct). It was also possible to classify between human and android images (see Figure \ref{fig:RobotsMVPA}d), however performance was much poorer, with a maximum of 62\% correct. Classification accuracy had an initial peak around 100ms that provided compelling evidence for above chance classification (BF\textgreater30), and a later region of above-chance classification between 500 and 700ms. In the Discussion we speculate that these two time periods might correspond to distinct types of signal associated with the uncanny valley.

Finally, we analysed the rating data from a set of 10 questionnaire items for 8 stimuli from each category. The results are summarised in Figure \ref{fig:robotratings}, and in general show differences between stimulus categories along most dimensions. Of particular note, the U-shaped function predicted by the uncanny valley effect was apparent for ratings along the Dislike-Like (Figure \ref{fig:robotratings}d), Unfriendly-Friendly (Figure \ref{fig:robotratings}e), and Anxious-Relaxed (Figure \ref{fig:robotratings}i) dimensions. These are all dimensions with emotional valence, indicating support for the `uncanniness' of our android stimuli. However we note that the android and robot categories were typically rated as being more similar to each other than to the real human faces. Following Experiment 1, we sought to generalise our results to a different stimulus set, and next report the results of Experiment 2 which used hyper-realistic silicone masks.

@@

\caption{Summary of response data and grand mean ERP for Experiment 2. Panel (a) shows d-prime scores for identifying images of human faces (blue), silicone masks (green) and Halloween masks (red). Small points show individual participants, and the larger symbols with error bars indicate the group mean and bootstrapped 95\% confidence intervals. Panel (b) plots reaction times in the same format (note the logarithmic x-axis). Panel (c) shows the grand mean ERP across all participants and conditions, pooled across electrodes P6, P8, PO6 and PO8 (see inset). The shaded region around the curve illustrates the 95\% confidence interval, and the grey rectangle at the foot indicates the stimulus duration.}\label{fig:MasksData}
\end{figure}

@@

\caption{Univariate and multivariate comparisons across image type for Experiment 2. Panel (a) shows the ERPs comparing human faces (blue) and Halloween masks (red), and panel (b) compares human faces (blue) and silicone masks (green). Panels (c) and (d) show multivariate pattern classification accuracy for the same comparisons. Points at y = -8 and y = 25 indicate Bayes factor scores for comparisons between ERPs (a,b) and comparing classification accuracy to chance (50\% correct; c,d).}\label{fig:MasksMVPA}
\end{figure}

@@

We subsequently obtained ratings from an independent sample of 20 participants using the same stimuli as in the EEG experiment. This time, we asked for explicit ratings of emotional expressiveness, realism, and uncanniness, as well as repeating the binary real face vs mask rating. Real faces were rated highest for emotional expressiveness (M=3.9) and realism (M=5.9), and lowest for uncanniness (M=2.7) (Figure \ref{fig:maskratings}a-c). The realistic silicone masks were rated highest for uncanniness (M=4.5), however this was not dramatically higher than for the Halloween masks (M=4.3). Arguably making judgements of uncanniness is less appropriate for masks that are not intended to be realistic, though our data do qualitatively conform to the U-shaped function expected by the uncanny valley hypothesis. The pattern of d-prime scores (Figure \ref{fig:maskratings}d) was similar to those obtained in the main experiment (Figure \ref{fig:MasksData}b), with generally higher scores attributable to the unlimited inspection time permitted in this online follow-up experiment

@@

Across two experiments using diverse stimuli, we identified a potential neurophysiological signature of the `uncanny valley' effect. EEG responses to androids or silicone masks could be distinguished from responses to human faces at around 100ms after stimulus onset, and also in a later time window around 500-800ms after stimulus onset. There were no clear differences in the unimodal ERP response at posterior electrodes, but performance of a multivariate pattern classifier was above chance in these time windows. This is a different pattern from that observed for more obviously non-human stimuli (robots and Halloween masks), where there were both univariate and multivariate differences, and the multivariate discrimination accuracy was above chance for an extended time window. Perceptual judgements indicated that identification performance for uncanny valley stimuli was relatively poor, indicating confusion with real human images. We also confirmed that android images were perceived more negatively than either human or robot images, and that silicone masks were perceived as more uncanny than human faces. The similarity in results across our two experiments is striking and constitutes an internal conceptual replication of our main findings.

@@

Our use of hyper-realistic silicone masks is novel in the context of the study of uncanny valley effects. Previous studies using these masks have demonstrated that they are difficult to distinguish from real faces (Sanders et al., 2017, 2019), including in applied settings such as simulated passport control (Robertson et al., 2020, 2024), and show large individual differences (Sanders \& Jenkins, 2018). Image analysis indicates that good identification performance for silicone masks is typically based on attention to the region below the eyes (Sanders \& Jenkins, 2018), however it is plausible that many observers are not explicitly aware of the cues they use to perform this judgement. This might contribute to both the early and late components identified in this study, and presumably also to the subjective sensation of `uncanniness' that is characteristic of the phenomenon.
